---
title: "ADNI datasets"
author: "Shahzad"
date: "`r format(Sys.Date(), '%A, %d %B %Y')`"
css: layout.css
output:
  html_document:
    number_sections: no
    code_folding: hide
    theme: united
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

<!-- Defining font size for the document -->

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F,
                      eval = T)
# load packages
pacman::p_load(data.table,
               here,
               gtsummary,
               flextable,
               rio,
               DT,
               UpSetR,
               tidyverse)
```
# List the Biomarker files in the folders 
```{r}
# Set your ADNI dataset path
data_path <- "/home/rstudio/ADNI_DATA_2025"
list.files(file.path(data_path, "Biospecimen"), full.names = TRUE)
list.files(file.path(data_path, "Enrollment"), full.names = TRUE)
biomarker_file <- file.path(data_path, "Biospecimen", "UPENNBIOMK_MASTER_FINAL_20May2023.csv")
diagnosis_file <- file.path(data_path, "Assessments", "DXSUM_PDXCONV_ADNIALL_19May2023.csv")

#biomarker_file <- file.path(data_path, "UPENN_PLASMA_FUJIREBIO_QUANTERIX_17Apr2025.csv")
#diagnosis_file <- file.path(data_path, "DXSUM_17Apr2025.csv")

# Load data
biomarkers <- read_csv(biomarker_file)
diagnosis <- read_csv(diagnosis_file)

```

```{r}
# Code for preprocessing used from published project https://doi.org/10.1093/brain/awae176
biomarkers <- subset(biomarkers, RUNDATE>=as.Date("2016-11-17"))

#go through each RID and compute median of values at duplicate time points
biomarkers <- biomarkers %>%
  mutate(across(c(ABETA42, TAU, PTAU), ~ as.double(gsub("<|>", "", .))))

# 
biomarkers <- biomarkers %>%
  mutate(RID_VISCODE2 = paste(RID, VISCODE2, sep = "_"))

biomarkers_clean <- biomarkers %>%
  group_by(RID_VISCODE2) %>%
  summarise(
    PHASE = first(PHASE),
    RID = first(RID),
    VISCODE2 = first(VISCODE2),
    EXAMDATE = first(EXAMDATE),
    ABETA40 = median(ABETA40, na.rm = TRUE),
    ABETA42 = median(ABETA42, na.rm = TRUE),
    TAU = median(TAU, na.rm = TRUE),
    PTAU = median(PTAU, na.rm = TRUE),
    .groups = "drop"
  )

```

# Read the biomarker files and calculate the AT categories for the biomarkers
```{r}
# Converting the variables to numeric 
biomarkers_clean$ABETA42<-as.numeric(biomarkers_clean$ABETA42)
biomarkers_clean$TAU<-as.numeric(biomarkers_clean$TAU)
biomarkers_clean$PTAU<-as.numeric(biomarkers_clean$PTAU)
# Creating an ATN category profiles of individuals 
biomarkers_clean$ATN_category2<-NA
biomarkers_clean$ATN_category2[biomarkers_clean$ABETA42>976.6 & biomarkers_clean$PTAU<21.8]<-'A_T_'
biomarkers_clean$ATN_category2[biomarkers_clean$ABETA42<976.6 & biomarkers_clean$PTAU<21.8]<-'A+T_'
biomarkers_clean$ATN_category2[biomarkers_clean$ABETA42<976.6 & biomarkers_clean$PTAU>21.8]<-'A+T+'

  
```
# Add the diagnosis information

```{r}

dxsum<-read_csv(file.path(data_path, "Assessments", "DXSUM_PDXCONV_ADNIALL_19May2023.csv"))

registry <- read_csv(file.path(data_path, "Enrollment", "REGISTRY_19May2023.csv"))
arm<-read_csv(file.path(data_path, "Enrollment", "ARM_19May2023.csv"))

# dxsum$DXCHANGE2 <- NA
# Assign DX CHANGE based on manual coding
# dxsum$DXCHANGE2[dxsum$DXCONV == 0 & dxsum$DXCURREN == 1] <- 1
# dxsum$DXCHANGE2[dxsum$DXCONV == 0 & dxsum$DXCURREN == 2] <- 2
# dxsum$DXCHANGE2[dxsum$DXCONV == 0 & dxsum$DXCURREN == 3] <- 3
# dxsum$DXCHANGE2[dxsum$DXCONV == 1 & dxsum$DXCONTYP == 1] <- 4
# dxsum$DXCHANGE2[dxsum$DXCONV == 1 & dxsum$DXCONTYP == 3] <- 5
# dxsum$DXCHANGE2[dxsum$DXCONV == 1 & dxsum$DXCONTYP == 2] <- 6
# dxsum$DXCHANGE2[dxsum$DXCONV == 2 & dxsum$DXREV == 1] <- 7
# dxsum$DXCHANGE2[dxsum$DXCONV == 2 & dxsum$DXREV == 2] <- 8
# dxsum$DXCHANGE2[dxsum$DXCONV == 2 & dxsum$DXREV == 3] <- 9

# Identify variable to keep for merged data
armVars <- c("RID","Phase","ARM","ENROLLED")
dxsumVars <- c("RID","Phase", "VISCODE2","DXCHANGE","VISCODE")
# merge data
dxarm <- merge(dxsum[dxsumVars], arm[armVars], by=c("RID","Phase"))
# baseline data
baseData <- dxarm[dxarm$VISCODE2=='bl' & dxarm$ENROLLED%in% c(1,2,3),]

# assign baseline diagnosis
baseData$baselineDx<-NA
baseData$baselineDx[(baseData$DXCHANGE %in% c(1,7,9)) & baseData$ARM != 11 ] = 1
baseData$baselineDx[(baseData$DXCHANGE %in% c(1,7,9)) & baseData$ARM == 11 ] = 2
baseData$baselineDx[(baseData$DXCHANGE %in% c(2,4,8)) & baseData$ARM == 10 ] = 3
baseData$baselineDx[(baseData$DXCHANGE %in% c(2,4,8)) & baseData$ARM != 10 ] = 4
baseData$baselineDx[(baseData$DXCHANGE %in% c(3,5,6))] = 5
# merge baseline diagnosis
baseVars <- c("RID","baselineDx")
dxarm <- merge( dxarm, baseData[baseVars],by=c("RID"))

# Merging with the registry files 
# identify variable to keep for merged data
regVars <-c("RID", "Phase", "VISCODE", "VISCODE2",
            "EXAMDATE", "PTSTATUS", "RGCONDCT", "RGSTATUS",
            "VISTYPE")
# merge data
dxarm_reg <-left_join(dxarm,registry[regVars], by=c("RID", "Phase", "VISCODE"))

# Merging the DXCHANGE AND baselineDx variables with the main df.
df<-dxarm_reg %>% 
  mutate(RID_VISCODE2=paste0(RID,'_',VISCODE2.x)) %>% 
  select(RID_VISCODE2,DXCHANGE,baselineDx) %>% 
  merge(biomarkers_clean,by='RID_VISCODE2',all.x =T)

```

# Add date to calculate the progression analysis

```{r}
# Upload the registry dataset 
registry <- read_csv(file.path(data_path, "Enrollment", "REGISTRY_19May2023.csv"))
registry_dates <- registry %>% mutate(RID_VISCODE2=paste0(RID,'_',VISCODE2)) %>% 
  select(RID_VISCODE2,Phase, RID, VISCODE2, EXAMDATE) 

# Merge with your df
df_time <- df %>% select(-EXAMDATE,-PHASE, -RID, -VISCODE2) %>% 
  left_join(registry_dates, by = "RID_VISCODE2")

# Baseline data 
baseline_dates <- df_time %>%
  filter(VISCODE2 == "bl") %>%
  select(RID, baseline_date = EXAMDATE)

# Merge date information with full dataset with dates 
df_days <- df_time %>%
  left_join(baseline_dates, by = "RID") %>%
  mutate(days_since_bl = as.numeric(difftime(EXAMDATE, baseline_date, units = "days")))
```

# Add APOE information
```{r}
adnimerg<-read_csv(file.path(data_path, "Study_Info", "ADNIMERGE_19May2023.csv"))
df<-adnimerg %>% 
  select(RID,VISCODE,APOE4,PTGENDER,AGE) %>% 
  mutate(RID_VISCODE2 = paste0(RID,'_',VISCODE)) %>% 
  select(-RID,-VISCODE) %>% 
  merge(., df_days,by='RID_VISCODE2', all.y = T)

```

# Creating the dataset for the survival analysis 
```{r}
######################################################################################################
#################################################### Progression analysis 
#################################################### 1 = Stable:NL to NL, (561)
#################################################### 2 = Stable:MCI to MCI, (809)
#################################################### 3 = Stable:AD to AD, (326)
#################################################### 4 = Conv:NL to MCI, (15)
#################################################### 5 = Conv:MCI to AD, (50)
#################################################### 6 = Conv:NL to AD, 7 = Rev:MCI to NL, (9) 8 = Rev:AD to MCI (3), 9 = Rev:AD to NL.
#################################################### Baseline diagnosis: as 1:Normal (495),2:SMC (72),3:EMCI (381),4:LMCI (575),5:AD (250),
# Only keep EMCI/LMCI at baseline
mci_baseline <- df %>%
  filter(VISCODE2 == "bl", baselineDx %in% c(3, 4)) %>%
  select(RID, baselineDx, ABETA42, TAU, PTAU, ATN_category2, baseline_date, APOE4, PTGENDER,AGE)

# Get time to AD conversion (DXCHANGE == 5)
conversion <- df %>%
  filter(DXCHANGE == 5, !is.na(days_since_bl)) %>%
  group_by(RID) %>%
  summarise(time_to_event = min(days_since_bl), event = 1)

censored <- df %>%
  filter(DXCHANGE == 2, !is.na(days_since_bl)) %>%
  group_by(RID) %>%
  summarise(time_to_event = max(days_since_bl), event = 0)

# Combine
surv_data <- bind_rows(conversion, censored) %>%
  left_join(mci_baseline, by = "RID") %>%
  mutate(time_to_event_months = time_to_event / 30.44)
```


# Survial analysis
```{r}
library(survival)
library(survminer)
fit <- survfit(Surv(time_to_event_months, event) ~ APOE4, data = surv_data)
ggsurvplot(fit,
           data = surv_data,
           risk.table = TRUE,
           pval = TRUE,
           xlab = "Time from Baseline (months)",
           ylab = "Survival Probability (MCI âž AD)",
           title = "Progression from MCI to AD by ATN Category",
           palette = "Dark2")
```
# Start preparting multi-omics dataset: 
## Add proteomics dataset
```{r}
library(janitor)
proteomics<-read.csv("/home/rstudio/new_direc/data_unziped/Biospecimen/BiomarkersConsortiumCSF_ADNI1_02Aug2013/Biomarkers Consortium ADNI CSF QC Multiplex data.csv")
Baseline_phenotype <- dxarm_reg %>%
  select(RID, Phase, VISCODE2.x, VISCODE, baselineDx) %>%
  mutate(RID_VISCODE2 = paste(RID, VISCODE2.x, sep = "_")) %>%
  filter(VISCODE2.x == "bl") %>%
  select(RID_VISCODE2, baselineDx) 

proteomics_with_id<-proteomics %>% select(-id, -sampid) %>% 
  mutate(RID_VISCODE2=paste(rid,visit_code, sep='_')) %>% 
  select(-rid,-visit_code)

# Replace all dots inside the matrix with NA.
proteomics_with_id[1:159] <- sapply(proteomics_with_id[1:159], function(x) ifelse(x == ".", NA, x))

# Calculate the number of missing proteins 
missing_summary <- sapply(proteomics_clean, function(x) sum(is.na(x)))
proteins_with_no_missing<-names(missing_summary)[which(missing_summary==0)]
proteomics_with_no_missing<-proteomics_with_id %>% select("RID_VISCODE2",all_of(proteins_with_no_missing))
# we have removed 76 proteins which were missing for all participants
# length(names(missing_summary)[which(missing_summary>1)]) and confirmed by apply(proteomics_clean, 1, function(x) sum(is.na(x)))

# Merge with the phenotype 
proteomics_with_phenotype<-proteomics_with_no_missing %>% left_join(Baseline_phenotype,by='RID_VISCODE2')
```

# Run some checks on the prepared dataset

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

proteomics_long <- proteomics_with_no_missing %>%
  pivot_longer(
    cols = -RID_VISCODE2, 
    names_to = "Marker",
    values_to = "Value"
  )

# Density plots for markers 
ggplot(proteomics_long, aes(x = Value)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  facet_wrap(~ Marker, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Proteomics", x = "Value", y = "Density")

# Boxplots for the proteomics data
ggplot(proteomics_long, aes(x = Marker, y = Value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```





